The Werewolf Among Us: Humans vs LLMs in Multi-Agent Games

### 2025-05-07 _tex/paper.tex
Built site for gh-pages

\title{The Werewolf Among Us: Humans vs LLMs in Multi-Agent Games}
\author{Bhavana Jonnalagadda \and Riley Jones}
\date{2025-05-06}
\begin{document}
\maketitle
\begin{abstract}
Abstract TODO
\end{abstract}

@@

\begin{itemize}
\item
  Social deduction games like \emph{Werewolf} offer a clear way to
  evaluate how agents deceive, persuade, and reason in group
  settings(\citeproc{ref-wikiwerewolf}{Wikipedia contributors 2024}). In
  these games, players have limited information, hidden identities, and
  must convince others while trying to figure out who is lying. These
  challenges closely match real life situations involving trust,
  negotiation, and manipulation.
\item
  We wanted to compare how humans and large language models (LLMs)
  handle these situations. To do this, we used two recent datasets:

  \begin{itemize}
  \tightlist
  \item
    \emph{Werewolf Among Us} (Lai et al., 2022), a collection of real
    human gameplay annotated with persuasion strategies,
  \item
    \emph{Werewolf Arena} (Bailis et al., 2024), a simulated environment
    where LLM agents play the game autonomously.
  \item
    While both studies show Werewolf generates complex strategic
    language, neither compares human and LLM behavior directly.
  \end{itemize}
\item
  Our project addresses this gap. We analyzed transcripts from both
  datasets, matched them by role, round, and persuasion strategy, and
  compared how humans and LLMs lie, persuade, and detect deception.
\item
  By annotating utterances with the same set of persuasive strategies,
  we clearly show how synthetic agents differ from or resemble humans
  when navigating deception in adversarial group interactions.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Human dataset description
  (\citeproc{ref-laiWerewolfUsMultimodal2022}{Lai et al. 2022})
\item
  Is specifically for a form of one-night werewolf

  \begin{itemize}
  \tightlist
  \item
    Describe key differences
  \end{itemize}
\item
  Used specifically for the text available

  \begin{itemize}
  \tightlist
  \item
    and annotations of persuasion strategy on the text
  \end{itemize}
\end{itemize}

\subsection{Werewolf Arena}\label{werewolf-arena}

\begin{itemize}
\tightlist
\item
  (\citeproc{ref-bailisWerewolfArenaCase2024}{Bailis, Friedhoff, and
  Chen 2024})
\item
  Discuss the framework, how it works, prompts, etc
\item
  Discuss what types of runs we did
\item
  Discuss the data included in output
\item
  Talk about how we had to annotate the LLM speech with persuasion
  strategies ourselves
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Formatted data to match, performed various comparisons
\end{itemize}

>>>>

\title{The Werewolf Among Us: Humans vs LLMs in Multi-Agent Games}
\author{Bhavana Jonnalagadda \and Riley Jones}
\date{2025-05-07}
\begin{document}
\maketitle
\begin{abstract}
Abstract TODO
\end{abstract}

@@

Social deduction games like Werewolf offer a clear way to evaluate how
agents deceive, persuade, and reason in group settings
(\citeproc{ref-wikiwerewolf}{Wikipedia contributors 2024}). In these
games, players operate with limited information and hidden identities,
attempting to convince others while trying to uncover deception
themselves. These dynamics mirror real-world challenges involving trust,
negotiation, and manipulation. To explore how humans and large language
models (LLMs) navigate such scenarios, we analyzed two key datasets:
Werewolf Among Us (\citeproc{ref-laiWerewolfUsMultimodal2022}{Lai et al.
2022}), which features real human gameplay annotated with persuasion
strategies, and Werewolf Arena
(\citeproc{ref-bailisWerewolfArenaCase2024}{Bailis, Friedhoff, and Chen
2024}), a simulated environment in which LLM agents autonomously play
the game. While both studies demonstrate that Werewolf elicits rich,
strategic language, neither directly compares human and LLM behavior.
Our project fills this gap. We analyzed transcripts from both datasets,
aligning them by role, round, and persuasive strategy, and compared how
humans and LLMs lie, persuade, and detect deception. By annotating all
utterances using a consistent taxonomy of persuasive strategies, we
expose key differences and similarities in how synthetic agents and
humans handle adversarial group interactions.

@@

We used the \emph{Werewolf Among Us} dataset {[}Lai et al.
(\citeproc{ref-laiWerewolfUsMultimodal2022}{2022}){]}, containing
annotated dialogues from over 150 real games of One Night Werewolf and
Avalon. These games differ from classic Werewolf by having only one
round of discussion and voting, not eliminating players during gameplay,
and featuring specialized roles beyond Villager and Werewolf. The
dataset includes detailed annotations of persuasion strategies for each
utterance, such as accusations, defenses, and identity claims. Our
analysis specifically utilized textual transcriptions and strategy
annotations for direct comparison.

\subsection{Werewolf Arena (LLM
Dataset)}\label{werewolf-arena-llm-dataset}

The \emph{Werewolf Arena} dataset {[}Bailis, Friedhoff, and Chen
(\citeproc{ref-bailisWerewolfArenaCase2024}{2024}){]} comprises
simulated classic Werewolf games played by autonomous LLM agents. Unlike
one-round human games, these simulations include multiple rounds
alternating between night (secret actions) and day (open discussion).
Each agent receives a role (Villager, Werewolf, Seer, Doctor) and
interacts through tailored prompts generated via an LLM API.

We conducted simulations using five LLM models: GPT-4o, GPT-4.1,
GPT-4o-mini, DeepSeek-Chat, and DeepSeek-Reasoner. Two configurations
were tested:

\begin{itemize}
\tightlist
\item
  8 players with 8 discussion rounds
\item
  10 players with 6 discussion rounds
\end{itemize}

We selected these settings to provide ample opportunity for villagers to
coordinate and demonstrate persuasive behaviors.

Agents participated in gameplay through a graphical user interface
(GUI), depicted in Figure Figure~\ref{fig-wwa-gui}. The GUI displays the
game state, including player roles, actions, and current discussion
rounds, enabling monitoring of the gameplay progression.

\textbf{Figure 1.} GUI of Werewolf Arena simulation environment.

\pandocbounded{\includegraphics[keepaspectratio]{Images/WWA_GUI.png}}

\caption{\label{fig-wwa-gui}GUI of Werewolf Arena simulation}

A central feature in Werewolf Arena is the dynamic turn-taking system
implemented via a bidding mechanism. Rather than a fixed speaking order,
agents bid for speaking turns based on urgency and strategic necessity,
closely simulating real-world group discussions. Bidding levels range
from passive observation to urgent direct responses:

\begin{itemize}
\tightlist
\item
  0: Observe quietly
\item
  1: Share general thoughts
\item
  2: Contribute critical and specific information
\item
  3: Urgent need to speak
\item
  4: Respond directly after being addressed or accused
\end{itemize}

The highest bidder speaks next, with ties broken by prioritizing agents
directly mentioned in preceding turns. This mechanism captures nuanced
strategic communication decisions made by agents throughout the game.

Agents utilize specialized prompts reflecting their current role, memory
state, and game context. The prompts guide strategic interactions,
influencing agent decisions in voting, debating, and night actions.
After generating dialogues through the LLM API, we manually annotated
these interactions using the persuasion strategy categories from the
human dataset.

@@

Annotations were standardized across both datasets for direct
comparative analysis. Our analyses explored frequency distributions of
persuasion strategies, role-based comparisons (villager vs.~werewolf),
and strategic differences between human and LLM-generated dialogues.
